---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# swaprinc

<!-- badges: start -->
<!-- badges: end -->

The goal of swaprinc is to help compare a regression model with raw variables to
a regression model where some of the raw variables have been swapped out for
principal components.

## Installation

You can install the development version of swaprinc from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mncube/swaprinc")
```

## A Simple Example

In the example below, a regression model is used to estimate the relation between
x1 and y after controlling for x2 - x5.  

Using the default engine of "stats" specifies that stats::lm will be used to fit 
the statistical model and using the default prc_eng of "stats" specifies that stats::prcomp will be used to extract principal components.  

The formula parameter specifies the "raw model" which will be passed to stats::lm.
The pca_vars and n_pca_components parameters specify that raw variables x2 - x5
will be used to extract 2 principal components, and then the following "pca model"
will be passed to stats::lm: y ~ x1 + PC1 + PC2

```{r simple}
library(swaprinc)

  # Create a small simulated dataset
  set.seed(40)
  n <- 50
  x1 <- rnorm(n)
  x2 <- rnorm(n, 5, 15)
  x3 <- rnorm(n, -5.5, 20)
  x4 <- rnorm(n, 3, 3)
  x5 <- rnorm(n, -2, 4)
  y <- 1 + 2 * x1 + 3 * x2 + 2.5*x4 - 3.5*x5 + rnorm(n)
  data <- data.frame(y, x1, x2, x3, x4, x5)

  # Run swaprinc with
  swaprinc_result <- swaprinc(data,
                              formula = "y ~ x1 + x2 + x3 + x4 + x5",
                              pca_vars = c("x2", "x3", "x4", "x5"),
                              n_pca_components = 2)
  
  # Summarize raw model
  summary(swaprinc_result$model_raw)
  
  # Summarize pca model
  summary(swaprinc_result$model_pca)
  
  # Get model comparisons
  print(swaprinc_result$comparison)

```
