---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# swaprinc

<!-- badges: start -->
<!-- badges: end -->

The objective of swaprinc is to streamline the comparison between a regression 
model using original variables and a model in which some of these variables have 
been swapped out for principal components.

## Installation

You can install the development version of swaprinc from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mncube/swaprinc")
```

## A Simple Example

In the simple example provided, a regression model estimates the relationship 
between x1 and y, while controlling for variables x2 through x10.

By using the default engine, "stats", the statistical model is fitted with 
stats::lm, and by using the default prc_eng, "stats", principal components are 
extracted with stats::prcomp.

The "raw model" is specified by the formula parameter, which is passed to stats::lm. 
The pca_vars and n_pca_components parameters indicate that variables x2 to x10 
will be used to extract three principal components. Subsequently, the "PCA model" 
is passed to stats::lm as follows: y ~ x1 + PC1 + PC2 + PC3.

By setting the lpca_center and lpca_scale parameters to 'pca', the data in pca_vars 
will be centered and scaled according to the guidelines in the 
[Step-by-Step PCA](https://cran.r-project.org/web/packages/LearnPCA/vignettes/Vig_03_Step_By_Step_PCA.pdf) vignette before being passed to stats::prcomp. The miss_handler parameter, 
set to 'omit', ensures that only complete cases are included by subsetting the 
data frame rows with stats::complete.cases.

```{r simple}
library(swaprinc)

  # Create a small simulated dataset
  set.seed(40)
  n <- 50
  x1 <- rnorm(n)
  x2 <- rnorm(n, 5, 15)
  x3 <- rnorm(n, -5.5, 20)
  x4 <- rnorm(n, 3, 3) + x3*1.5
  x5 <- rnorm(n, -2, 4) + x3*.25
  x6 <- rnorm(n, -5, 5) + x4
  x7 <- rnorm(n, -2, 6)
  x8 <- rnorm(n, 2, 7)
  x9 <- rnorm(n, -2, 3) +x2*.4
  x10 <- rnorm(n, 5, 4)
  y <- 1 + 2 * x1 + 3 * x2 + 2.5*x4 - 3.5*x5 + 2*x6 + 1.5*x7 + x8 + 2*x9 + x10 + rnorm(n)
  data <- data.frame(y, x1, x2, x3, x4, x5, x6, x7, x8, x9, x10)

  # Run swaprinc with
  swaprinc_result <- swaprinc(data,
                              formula = "y ~ x1 + x2 + x3 + x4 + x5 + x5 + x6 + x7 + x8 + x9 + x10",
                              pca_vars = c("x2", "x3", "x4", "x5", "x6", "x7", "x8", "x9", "x10"),
                              n_pca_components = 3,
                              lpca_center = "pca", 
                              lpca_scale = "pca",
                              miss_handler = "omit")
  
  # Summarize raw model
  summary(swaprinc_result$model_raw)
  
  # Summarize pca model
  summary(swaprinc_result$model_pca)
  
  # Get model comparisons
  print(swaprinc_result$comparison)

```


## The Motivating Example

A common challenge in applied statistics and data science involves performing 
logistic regression with a set of categorical independent variables. In this 
motivating example, swaprinc is employed to compare a 'raw' logistic regression 
model containing seven categorical independent variables with a 'pca' logistic 
regression model. The latter model replaces six of the independent variables with 
their first three principal components, using Gifi::princals to extract principal
components. For a comprehensive tutorial on Gifi, refer to 
[Nonlinear Principal Components Analysis: Multivariate Analysis with Optimal Scaling (MVAOS)](https://www.css.cornell.edu/faculty/dgr2/_static/files/R_html/NonlinearPCA.html#2_Package).

I recommend using the 'broom' and 'broom.mixed' packages to summarize model 
results when utilizing the '*_options' parameters for passing arguments to 
functions within 'swaprinc'. This approach helps prevent 
[overly extensive summaries caused by 'do.call'](https://stackoverflow.com/questions/75512192/r-do-call-function-returns-to-much/75512429#75512429).

```{r motivation}
 # Create a small simulated dataset
  set.seed(42)
  n <- 50
  x1 <- rnorm(n, 0.5, 4)
  x2 <- rnorm(n, 3, 15)
  x3 <- rnorm(n, -2.5, 5)
  x4 <- -2.5*x2 + 3*x3 + rnorm(n, 0, 4)
  x5 <- x2*x3 + rnorm(n, -5, 5)*rnorm(n, 5, 10)
  x6 <- rnorm(n, -2, 4)*rnorm(n, 3, 5)
  x7 <- x4 + x6 + rnorm(n, 0, 3)
  y <- 1 + 2*x1 + 3*x2 + -2*x3 + .5*x4 + x5 + 1.5*x6 + x7 + rnorm(n)
  data <- data.frame(y, x1, x2, x3, x4, x5, x6, x7)

  # Categorize the variables
  yq <- stats::quantile(data$y,c(0,1/2, 1))
  x1q <- stats::quantile(data$x1,c(0,1/2, 1))
  x2q <- stats::quantile(data$x2,c(0,1/4,3/4,1))
  x3q <- stats::quantile(data$x3,c(0,2/5,3/5,1))
  x4q <- stats::quantile(data$x4,c(0,1/5,4/5,1))
  x5q <- stats::quantile(data$x5,c(0,2/5,3/5,1))
  x6q <- stats::quantile(data$x6,c(0,2/5,4/5,1))
  x7q <- stats::quantile(data$x7,c(0,2/5,3/5,1))


  data <- data %>% dplyr::mutate(
    y = cut(y, breaks=yq, labels=c("0", "1"),include.lowest = TRUE),
    x1 = cut(x1, breaks=x1q, labels=c("control", "treatment"),include.lowest = TRUE),
    x2 = cut(x2, breaks=x2q, labels=c("small","medium","large"),include.lowest = TRUE),
    x3 = cut(x3, breaks=x3q, labels=c("short","average","tall"),include.lowest = TRUE),
    x4 = cut(x4, breaks=x4q, labels=c("lowbit","most","highbit"),include.lowest = TRUE),
    x5 = cut(x5, breaks=x5q, labels=c("under","healthy","over"),include.lowest = TRUE),
    x6 = cut(x6, breaks=x6q, labels=c("small","medium","large"),include.lowest = TRUE),
    x7 = cut(x7, breaks=x7q, labels=c("small","medium","large"),include.lowest = TRUE)) %>%
    dplyr::mutate(y = as.numeric(ifelse(y == "0", 0, 1)))

  # Run swaprinc with prc_eng set to Gifi
  swaprinc_result <- swaprinc(data,
                              formula = "y ~ x1 + x2 + x3 + x4 + x5 + x6 + x7",
                              pca_vars = c("x2", "x3", "x4", "x5", "x6", "x7"),
                              n_pca_components = 3,
                              prc_eng = "Gifi",
                              model_options = list(family = binomial(link = "logit")))
  
  # Summarize raw model
  broom::tidy(swaprinc_result$model_raw)
  
  # Summarize pca model
  broom::tidy(swaprinc_result$model_pca)
  
  # Get model comparisons
  print(swaprinc_result$comparison)
```


## Compare Multiple Models

Utilizing the same dataset as in the logistic regression model mentioned earlier, 
it is beneficial to compare outcomes for various swaps. In the example below, 
the compswap helper function facilitates the comparison of results with 2, 3, 4, 
and 5 principal components replacing six original independent variables. 

```{r compswap}
  # Run swaprinc with prc_eng set to Gifi
  compswap_results <- compswap(data,
                              formula = "y ~ x1 + x2 + x3 + x4 + x5 + x6 + x7",
                              .pca_varlist = list(c("x2", "x3", "x4", "x5", "x6", "x7")),
                              .n_pca_list = list(2, 3, 4, 5),
                              .prc_eng_list = list("Gifi"),
                              .model_options_list = list(list(family = binomial(link = "logit"))))

  # Show available models
  summary(compswap_results$all_models)
  
  # Get model comparisons
  print(compswap_results$all_comparisons)
  
  # View model summaries
  lapply(compswap_results$all_models, broom::tidy)
```
